Methods to prioritise pop-up active transport infrastructure

### 2020-06-13 article.Rmd

@@

At a time of reduced travel, fewer lanes dedicated to motor traffic are needed.
Based on this observation, we defined roads with a 'spare lane' as those on which there is more than one lane in either direction.
This definition assumes no alteration of the navigable network for motor vehicles (making two-way streets one-way is another option not explored in this analysis).

To identify road sections on which there is a spare lane we developed a simple algorithm that takes the OSM variable [`lanes`](https://wiki.openstreetmap.org/wiki/Key:lanes) if it is present and, if not, derives the number from the highway type and presence/absence of bus lanes.
All segments defined as having a spare lane using this method are shown in Figure \@ref(fig:levels) (left).
In future, this methodology could be enhanced to take into account the effect of lane widths, which increase the effective available space, and the presence of vehicle parking bays, which reduce it.

The filtering and grouping of road segments is shown in Figure \@ref(fig:levels). 
Segments are grouped with a 100m buffer, using the `igraph` R package; they are also filtered to exclude sections below a minimum length and cycling potential. 
The threshold length and cycling potential are adaptable depending on the nature of the region being studied and local cycling levels.
We can see that grouping the segments first then filtering based on mean group-level attributes results in a more cohesive network than filtering individual segments then grouping the results.

@@

>>>>

## Road attributes

At a time of reduced travel, fewer lanes dedicated to motor traffic are needed.
Based on this observation, we defined roads with a 'spare lane' as those on which there is more than one lane in either direction.
This definition assumes no alteration of the navigable network for motor vehicles (other options such as making two-way streets one-way are not explored in this analysis).

To identify road sections with a spare lane we developed a simple algorithm that takes the OSM variable [`lanes`](https://wiki.openstreetmap.org/wiki/Key:lanes) if it is present and, if not, derives the number from the highway type and presence/absence of bus lanes.
All segments defined as having a spare lane using this method are shown in Figure \@ref(fig:levels) (left).

We also included estimates of road width, flagging roads with mean estimated width >= 10m using data previously generated for the Cycling Infrastructure Prioritisation Tool.
We chose 10m because it represents the minimum typical width required to enable cycleway creation through road space reallocation.

@@

To ensure our route recommendations could achieve sufficient coherency, we undertook several stages of road segment filtering and grouping. 
Firstly, segments were grouped by road reference number (i.e. 'A' or 'B' road number).
Road segments with a reference number were grouped together with neighbours within a 100m buffer, using the `igraph` R package. 
Filtering then removed groups without distance weighted mean width >= 10 m or spare lanes along the majority of their length, and groups with distance weighted mean cycling potential below a minimum threshold defined as one twenty-fifth of the 99th percentile segment level cycling potential within the city.

Segments without a reference number were subjected to stricter filtering criteria, to prevent the inclusion of unwanted short segments on side streets. 
Any of these segments that had cycling potential below 30 were excluded from the analysis. 
The segments were then grouped using a 20m buffer. 
Filtering followed the same criteria as for other roads, plus an additional filter to remove groups with length below 300 m.

For all segments, a novel round of grouping (ignoring previous groups) with a 100m buffer was then used to remove groups with length below 500 m. 
This step removed short sections distant from any others, thus improving the coherency of the results.
Finally, road names were used to identify continuous road sections with the same name of length >= 500m.
Groups containing five or more different named roads were labelled "Unnamed road."

An example of the impact of grouping strategy is shown in Figure \@ref(fig:levels). 
Segments are grouped with a 100m buffer, using the `igraph` R package; they are also filtered to exclude sections below a minimum length and cycling potential. 
The threshold length and cycling potential are adaptable depending on the nature of the region being studied and local cycling levels.
We can see that grouping the segments first then filtering based on mean group-level attributes results in a more cohesive network than filtering individual segments then grouping the results.

@@

## Selection of top routes

Top routes were selected from the results of the previous steps. 
These must not be labelled "Unnamed road" or have existing cycleways along more than 80% of their length. 
A high threshold was chosen here because the presence of an existing cycleway on OSM does not mean that this is necessarily a high quality cycleway. 
Continuity of cycle provision is important for creating high quality networks [@parkin_designing_2018].
